## Functions and Conditionals

### Conditionals

The basic syntax for an if statement goes like this:

```{r}
if (logical statement) {
  do something
} else {
  do something else
}
```

The logical statements should boil down to a boolean; for example:

```{r}
a <- 10

if (a <= 5) {
  print(a)
} else {
  print("it's more than 5")
}
```

We can also add else if statements in the middle if we need more options.

```{r}
a <- 15

if (a <= 5) {
  print("small")
} else if (a > 5 & a <=10) {
  print("medium")
} else if (a > 10 | a == 7) {
  print("large")
} else {
  print("huge")
}
```

Neat! We can also nest conditionals just like we do in Python.

```{r}
a <-  5
if (a / 2 != 0) {
  if (a == 5){
    print("a is 5")
  }
}
```

Neat! Now, let's look at another core part of R programming - Loops!

### Loops

```{r}
for (value in 1:5){
  # Paste is a concatenation function
  print(paste(value, "is greater than", value - 1))
}
```

We can also do nifty stuff with our vectors, such as apply changes to specific indices of our data.

```{r}
a_vector <- c(1,1,1,1,1,1,1)

# change the first, third, and sixth elements to the value 6

for (i in c(1,3,6)){
  a_vector[i] <- 6
}

a_vector
```

We can loop over more than just lists:

```{r}
mat <- matrix(1, nrow = 2, ncol = 2)

for (row in 1:2){
  for (col in 1:2){
    mat[row,col] <- 0
    print(mat)
  }
}
```

Like Python, we have while loops in R:

```{r}
b <-  0

while (b <10){
  # increase b by 2
  b <- b + 2
  # print b
  print(b)
}
```

## Functions

When should we use functions?

```{r}
our_function <- function(optional_params){
  # do some stuff
}

```

An example:

```{r}
# Start with a vector of temperatures 

test_temp <- c(16.550540, 26.942005, 32.591088, 8.580116, 28.003873, 28.542391, 20.976820, 21.173577, 21.048836, 18.769735)

# Then we want a function that converts from C to F:
c_to_f <-  function(dat){
  temp_fahr <- dat*(9/5) + 32
  
  return(temp_fahr)
}

print(c_to_f(26))
```

This is all cool.. but it didn't do anything. We need to call the function!

```{r}
data_fahr <- c_to_f(test_temp)
data_fahr
```

Notice that in the background here, we are looping over each item in the array, but we never write a for loop! Keep this in mind, since when you pass in a large data set, the changes made by functions aren't quite immediate - the computer does have to go over each item in your data set, and change them one at a time.

A shortcut of sorts is to use the `apply()` function, which operates very similarly to Python's `map()`. Note that `apply` syntax goes as follows: `apply(X, MARGIN, FUN)` where X is the data set to use, MARGIN is specification for rows or columns, and FUN is the function to apply.

```{r}
test_mat <- matrix(c(1,1,1,2,2,2), nrow <- 2)
# 1 is for rows, 2 is for columns 
apply(test_mat, 1, sum)
```

It is common to use `mean`, `median`, `sum`, `min`, or `max` with `apply`, but we can also use our user-defined functions as well. Note that apply() only works with matrices.

```{r}
test_temp_mat <- matrix(c(16.550540, 26.942005, 32.591088, 8.580116, 28.003873, 28.542391, 20.976820, 21.173577, 21.048836, 18.769735), nrow = 2)

apply(test_temp_mat, 1, c_to_f)
```

### Homework

1.  Do 5 CodeWars problems in R. Share your solutions here:

    ```{r}
    # Write a function `greet` that returns "hello world!"
    greet <- function(){
      return("Hello world!")
    }

    message <- greet()
    print(message)
    ```

    ```{r}
    # We need a function that can transform a number (integer) into a string

    number_to_string <- function(num){
      return(as.character(num))
    }

    number_to_string(31)
    ```

    ```{r}
    # Create a function that takes an integer as an argument and returns "Even" for even numbers or "Odd" for odd numbers.

    evenOrOdd <- function(number){
      if (number %% 2 == 0){
        return("Even")
      } else{
        return("Odd")
      }
    }

    result <- evenOrOdd(4)
    result
    ```

    ```{r}
    # It's pretty straightforward. Your goal is to create a function that removes the first and last characters of a string. You're given one parameter, the original string. You don't have to worry with strings with less than two characters.

    remove_char <- function(original_string) {
      if(nchar(original_string) < 2) {
        return(original_string)
      } else {
        return(substr(original_string, 2, nchar(original_string) -1 ))
      }
    }

    result <- remove_char("banana")
    result
    ```

    ```{r}
    # In this kata, your job is to return the two distinct highest values in a list. If there're less than 2 unique values, return as many of them, as possible. The result should also be ordered from highest to lowest.

    two_highest <- function(input_vector) {
      unique_values <- sort(unique(input_vector), decreasing = TRUE)
      if (length(unique_values) < 2) {
        return(unique_values)
      } else {
        return(unique_values[1:2])
      }
    }

    result <- two_highest(c(15,20,25,30,35,21,1,42,4,54))
    result
    ```

2.  reate a function that checks an input of type `int` and returns fizz if the number is a multiple of 3, buzz if the number is a multiple of 5, and fizzbuzz if the number is a multiple of 3 and 5.

    ```{r}
    fizzbuzz <- function(x) {
      if (x %% 3 == 0 && x %% 5 == 0 ) {
        return ("fizzbuzz")
      } else if (x %% 5 == 0){
        return("buzz")
      } else if (x %% 3 == 0){
        return("fizz")
      } else {
        return("This number is not a multiple of 3 or 5")
      }
    }

    values <- c(1:20)

    for (value in values){
      result <- fizzbuzz(value)
      cat(paste(value, "-", result), "\n")
    }
    ```

3.  **Mario Challenge:** In the classic Super Mario games, Mario ends each level by ascending a flight of stairs and leaping off the top to arrive at the castle at the end. (See [here](https://youtu.be/-avspZlbOWU?t=55) for details). We can imagine that each of the bricks of the stairs is a character X. Create a function that prints out a set of stairs. For example:

    If the function is called marioStairs() and we pass in the value 4, we would expect to see stairs with a length of 4 and a height of 4:

    X

    XX

    XXX

    XXXX

    ```{r}
    # You may feel more comfortable figuring this out in Python first, then translating into R later.

    marioStairs <- function(length, height) {
      for (i in 1:height){
        spaces <- length - i
        
        cat(rep(' ', spaces), rep('X', i), '\n')
      }
    }

    marioStairs(4,4)
    ```
